Implementing Distributed Systems Directly with I/O Automata in Ioa++

### 2011-09-15 papers/paper1/component_model.tex
Revision.

Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then applying the effect if the precondition is true.
In general the scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if the precondition is true.
Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions, using the value produced by the output action subject to the precondition.
The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.

@@

In ioa++, recursive encapsulation means that an I/O automaton may contain and use other I/O automata.
Recursive encapsulation implies that a parent automaton is the result of composing its children and an anonymous \emph{kernel automaton}.
The kernel automaton allows a parent automaton to have its own state variables and actions.
The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.

Figure~\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles, e.g., Processor.
Named automata are labeled by a box in their upper left corner, e.g., C.
The sample event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
The Processor automaton is the root and C and D are its children.

@@

\paragraph{Parameters.}
Actions in the I/O automata model can defined using parameters.
Parameters are useful for situations requiring fan-in or a session.
For example, consider an automaton with an input actions $in?(s)$ that is to be bound to $N$ different output actions.
The solution is to parameterize the input action with a parameter that indicates the associated output actions, e.g., $in?[i](s)$ where $i$ takes on the integer values from 1 to $N$.
\emph{Parameterized} actions take a parameter while \emph{unparameterized} actions do not.
Parameters, for actions that require them, are specified in the binding and serve to identify the action.

@@

\paragraph{Example.}
Figure~\ref{example1} depicts a remote sensor systems as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
The Remote Sensor automaton consists of the Control automaton $Z$ and three Channel automata $A$, $B$, and $C$ representing three network connections.
The Control automaton $Z$ consists of the Sensor automaton $S$, the Filter automaton $F$, and the Statistics automaton $T$.
The Control automaton $C$ starts the sampling process with the uv-up output $Z.trigger!$.
The uv-up input $S.trigger?$ is executed atomically with $Z.trigger!$ due to the binding $(Z.trigger!, S.trigger?)$.
When the sampling process is complete, the v-up output $S.sample!(s)$ distributes the sample to the Filter $F$ and Statistics component $T$ via the $(S.sample!(s), F.in?(s))$ and $(S.sample!(s), T.in?(s))$ bindings.
Note that $S.sample!(s)$, $F.in?(s)$, and $T.in?(s)$ all agree on the type of the sample $s$.
The Filter automaton $F$ passes the filtered sample to the Control automaton $Z$ via the $F.out!(t) \to Z.sample?(t)$ binding.
The Statistics automaton $T$ calculates the statistics incrementally using the up internal $T.compute$.
The statistics can be polled by the Control automaton $Z$ using the bindings $(Z.request!, T.request?)$ and $(T.response!(u), Z.response?(u))$.
The Control automaton $Z$ contains a vp-output $Z.send![](m)$ and a vp-input $Z.recv?[](m)$ for sending and receiving network messages using the Channel automata $A$, $B$, and $C$.
The parameters 0, 1, and 2, are used to route messages to Channel automata $A$, $B$, and $C$ respectively using the session idea mentioned previously.
Opportunities for concurrent execution exist for the remote sensor depicted in figure~\ref{example1}.
The set of automata implied by $Z.trigger!$ is $\{Z, S\}$ while the set of automata implied by $T.compute$ is $\{T\}$.
Since the sets are disjoint, the two actions can be executed concurrently.

@@

\begin{figure}
\center
\includegraphics[width=\textwidth]{system_action}
\caption{System actions.
  An automaton contains output actions for creating, binding, unbinding, and destroying.
  The system automaton contains parameterized input actions for receiving the requests to create, binding, unbind, and destroy.
  The system automaton returns a result using the appropriate parameterized output action.
  The automaton contains input actions to receive the results.}
\label{system_action}
\end{figure}

\paragraph{The system automaton.}
Let the set of automata $A$ and set of bindings $B$ be managed by the \emph{system automaton} which represents the run-time system.
Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
The system automaton, in turn, has input actions for receiving requests for system actions, possibly some internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
Figure~\ref{system_action} shows the relationship between an automaton and the system automaton.

>>>>

Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then applying the effect if the precondition is true.
In general the scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if the precondition is true.
Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions,  subject to the precondition using the value produced by the output action.
The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.

@@

Applying recursive encapsulation to I/O automata means that an I/O automaton may contain and use other I/O automata.
Recursive encapsulation implies that a parent automaton is the result of composing its children and an anonymous \emph{kernel automaton}.
The kernel automaton allows a parent automaton to have its own state variables and actions.
The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.

Figure~\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles.
The type of the automaton appears in a box in the upper left or lower left corner, e.g., Processor.
The name of a named automata appears after its type, e.g., Converter C.
The sample event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
The Processor automaton is the root and C and D are its children.

@@

\paragraph{Parameters.}
Actions in the I/O automata model can be defined using parameters.
Parameters are useful for situations requiring fan-in or session semantics.
For example, consider an automaton with an input action $in?(s)$ that is to be bound to $N$ different output actions.
To get around the limitation that an input can only be bound to one output, we augment the input action with a parameter that indicates the associated output action, e.g., $in?[i](s)$ where $i$ takes on the integer values from 1 to $N$.
\emph{Parameterized} actions take a parameter while \emph{unparameterized} actions do not.
Parameters, for actions that require them, are specified in the binding and serve to identify the action.

@@

\paragraph{Example.}
Figure~\ref{example1} depicts a remote sensor systems as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
The Remote Sensor automaton consists of the Control automaton $Z$ and three Channel automata $A$, $B$, and $C$ representing three network connections.
The Control automaton $Z$ consists of the Sensor automaton $S$, the Filter automaton $F$, and the Statistics automaton $T$.
The Control automaton $C$ starts the sampling process with the uv-up output $Z.trigger!$.
The uv-up input $S.trigger?$ is executed atomically with $Z.trigger!$ due to the binding $(Z.trigger!, S.trigger?)$.
When the sampling process is complete, the v-up output $S.sample!(s)$ distributes the sample to the Filter $F$ and Statistics component $T$ via the $(S.sample!(s), F.in?(s))$ and $(S.sample!(s), T.in?(s))$ bindings.
Note that $S.sample!(s)$, $F.in?(s)$, and $T.in?(s)$ all agree on the type of the sample $s$.
The Filter automaton $F$ passes the filtered sample to the Control automaton $Z$ via the $(F.out!(t), Z.sample?(t))$ binding.
The Statistics automaton $T$ calculates the statistics incrementally using the up internal $T.compute$.
The statistics can be polled by the Control automaton $Z$ using the bindings $(Z.request!, T.request?)$ and $(T.response!(u), Z.response?(u))$.
The Control automaton $Z$ contains a vp-output $Z.send![](m)$ and a vp-input $Z.recv?[](m)$ for sending and receiving network messages using the Channel automata $A$, $B$, and $C$.
The parameters 0, 1, and 2, are used to route messages to Channel automata $A$, $B$, and $C$ respectively using the session idea mentioned previously.
Opportunities for concurrent execution exist for the remote sensor depicted in figure~\ref{example1}.
The set of automata implied by $Z.trigger!$ is $\{Z, S\}$ while the set of automata implied by $T.compute$ is $\{T\}$.
Since the sets are disjoint, the two actions can be executed concurrently.

@@

\begin{figure}
\center
\includegraphics[width=\textwidth]{system_action}
\caption{System actions.
  A base automaton contains output actions for creating, binding, unbinding, and destroying.
  The system automaton contains parameterized input actions for receiving the requests to create, binding, unbind, and destroy.
  The system automaton returns a result using the appropriate parameterized output action.
  The base automaton contains input actions to receive the results.}
\label{system_action}
\end{figure}

\paragraph{The system automaton.}
Let the set of automata $A$ and set of bindings $B$ be managed by the \emph{system automaton} which represents the run-time system.
Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
The system automaton, in turn, has input actions for receiving requests for system actions, possibly some internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
Figure~\ref{system_action} shows the relationship between a base automaton and the system automaton.

### 2011-09-15 papers/paper1/design.tex
Revision.

An implementation of the component model described in section~\ref{component_model} must provide a service for executing local actions and facilities for manipulating the set of automata and bindings.
The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
The Scheduler has the resposibility of selecting and executing local actions subject to the I/O automata model and concurrency constraints of~\ref{component_model}.
All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
The services provided by the base class are indicated by the Automaton layer of Figure~\ref{framework_architecture}.
The Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
In most situations, the services provided by the Automaton base class are too low-level to be used directly by a User Automaton.
An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Automaton base class.
Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.
We assume a POSIX environment whose native concurrency mechanism is threads.

\subsection{Scheduler}

@@

\paragraph{The Model.}
The model enforces the execution and concurrency constraints of section~\ref{component_model}.
Input actions are executed with output actions according to the current set of bindings.
The model allows executing actions concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
Since concurrent access to the model is allowed, the module implementing the model must thread-safe.

Our implementation of the model uses a two-level locking scheme to enforce the concurrent execution constraints.
Any action that modifies the set of automata and bindings must first acquire a write lock to prevent concurrent access.
All other actions acquire a read lock to implement single-writer/multiple reader semantics.
The second phase of executing a non-modifying action involves acquiring locks for the set of automata implied by an action.
The locks are acquired in a fixed order to prevent deadlock.
Once the locks are acquired, the precondition is evaluated, the local action and all bound input actions are executed, and the locks are released.

@@

The approach used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
The controller then has the responsibility of processing the set of requested actions according to some fair policy.
Let an action be \emph{enabled} if its precondition is true and \emph{disabled} otherwise.
Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
Consequently, an automaton is given a chance to declare actions to the scheduler after one of its actions is executed, i.e., changes state.
Since the state and therefore preconditions of all automata are independent, each automaton is responsible for scheduling its own actions.
For bootstrapping, each automaton is allowed to schedule actions when it is created.

@@

In contrast, an automaton should ensure that the scheduler will consider all actions that are currently enabled.
A simple technique for accomplishing this is to schedule all enabled actions.
More efficient schemes are possible if one considers which actions enable each other and which actions have already been scheduled.
Not declaring an enabled action violates the fairness of the controller since it will never select an action that may cause a state change.
Not scheduling an (enabled) action is a common mistake.

\paragraph{Binding predicates and scheduling.}
The preceding discussion on scheduling claims that an action in one automaton cannot affect preconditions in another automaton.
This is not entirely correct when one considers the system automaton and binding predicates.
Imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
When the automaton is created, its only action is disabled.
Consequently, it will not schedule its action and therefore not be executed.
Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
The bind changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound; a situation that occurs quite often in practice.
To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after a bind or unbind.

\paragraph{Actions of non-existent automata.}
Every action in the set of actions considered by the controller was scheduled by its associated automaton.
Since automata can be destroyed, an action in the set of actions owned by the controller might belong to an automaton that no longer exists.
Our solution is to check if the automaton exists before executing an action.
This check is performed by the model and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
Our solution assumes that each automaton is assigned a unique identifier and that the set of actions turns over comparatively faster than the set of identifiers.

@@

\paragraph{File descriptor events.}
File descriptors are used to communicate with the outside world in a POSIX environment.
Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects adversely affects the asynchronous and concurrent nature of I/O automata.
Thus, we require techniques that allow a program (1) to wait until a file descriptor is ready for I/O and (2) to perform non-blocking I/O operations.
This implies the use of either the Proactor or Reactor pattern.
For our framework, we used the Reactor pattern since (1) the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less standardized.

@@

\subsection{System Actions}

@@

\paragraph{Automaton and Binding managers.}
The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple applications but becomes unwieldy when creating complicated constellations.
The Automaton Managers and Binding Managers provide an individualized facade to the system action state machines.
Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
Automaton Managers can be observed using the Observer pattern to determine when and how the creation/destruction process succeeds.
Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created, thus, Binding Managers observe Automaton Managers.

@@

>>>>

An implementation of the component model described in section~\ref{component_model} must provide a service for executing local actions and facilities for manipulating the set of automata and bindings.
The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
The Scheduler has the resposibility of selecting and executing local actions subject to the I/O automata model and concurrency constraints of~\ref{component_model}.
All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
The services provided by the base class are indicated by the Base Automaton layer of Figure~\ref{framework_architecture}.
The Base Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
In most situations, the services provided by the Base Automaton class are too low-level to be used directly by a User Automaton.
An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Base Automaton class.
Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.
We assume a POSIX environment whose native concurrency mechanism is threads.

\subsection{Scheduler\label{scheduling}}

@@

\paragraph{The Model.}
The model enforces the execution and concurrency constraints of section~\ref{component_model}.
Input actions are executed with output actions according to the current set of bindings.
The model allows executing actions concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
Since concurrent access to the model is allowed, the module implementing the model must be thread-safe.

Our implementation of the model uses a two-level locking scheme to enforce the concurrent execution constraints.
Any action that modifies the set of automata and bindings must first acquire a write lock to prevent concurrent access.
All other actions acquire a read lock to implement single-writer/multiple reader semantics.
The second phase of executing a non-modifying action involves acquiring locks for the set of automata implied by an action.
The locks are acquired in a fixed order to prevent deadlock~\cite{havender1968avoiding}.
Once the locks are acquired, the precondition is evaluated, the local action and all bound input actions are executed, and the locks are released.

@@

The approach used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
The controller then has the responsibility of processing the set of requested actions according to some fair policy.
Let an action be \emph{enabled} if its precondition is true and \emph{disabled} otherwise.
Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
Consequently, an automaton is given a opportunity to declare actions to the scheduler after one of its actions is executed.
Since the state and therefore preconditions of all automata are independent, each automaton is responsible for scheduling its own actions.
For bootstrapping, each automaton is allowed to schedule actions when it is created.

@@

In contrast, an automaton should ensure that the scheduler will consider all actions that are currently enabled.
A simple technique for accomplishing this is to schedule all enabled actions.
More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
Not declaring an enabled action violates the fairness of the controller since it will never select an action that may cause a state change.
Not scheduling an (enabled) action is a common mistake.

\paragraph{Binding predicates and scheduling.}
The preceding discussion on scheduling claims that an action in one automaton cannot affect preconditions in another automaton.
This is not entirely correct when one considers the system automaton and binding predicates.
Imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
When the automaton is created, its only action is disabled.
Consequently, it will not schedule its action and therefore not be executed.
Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
The bind changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound; a situation that occurs quite often in practice.
To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after a successful bind or unbind.

\paragraph{Actions of non-existent automata.}
Every action in the set of actions considered by the controller was scheduled by its associated automaton.
Since automata can be destroyed, an action in the set of actions owned by the controller might belong to an automaton that no longer exists.
Our solution is to check if the automaton exists before executing an action.
This check is performed by the model and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
Our solution assigns each automaton a unique identifier called an \emph{automaton id (aid)} and that the set of actions in the scheduler turns over comparatively faster than the set of aids.

@@

\paragraph{File descriptor events.}
File descriptors are used to communicate with the outside world in a POSIX environment.
Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects adversely affects the asynchronous and concurrent nature of I/O automata.
Thus, we require techniques that allow a program (1) to wait until a file descriptor is ready for I/O and (2) to perform non-blocking I/O operations.
This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
For our framework, we used the Reactor pattern since (1) the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less standardized.

@@

\subsection{System Actions\label{system_action_section}}

@@

\paragraph{Automaton and Binding managers.}
The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple applications but becomes unwieldy when creating complicated constellations.
The Automaton Managers and Binding Managers provide an individualized facade to the system action state machines.
Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
Automaton Managers can be observed using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created, thus, Binding Managers observe Automaton Managers.

@@

\subsection{Implementation}

Ioa++ is implemented in C++ and available at \url{http://jrwilson.github.com/ioa/}.

### 2011-09-15 papers/paper1/programming_model.tex
Revision.

\paragraph{Declaring an automaton.}
All automata inherit from the common base class \verb+ioa::automaton+ which implements the system action interface.
The following code declares the \verb+asynch_lcr_automaton+:
\begin{lstlisting}
template <typename UID>
class asynch_lcr_automaton :
  public virtual ioa::automaton
{
  ...
};
\end{lstlisting}
The \verb+asynch_lcr_automaton+ is declared as a template where the \verb+UID+ template parameter must model the unique identifier concept.
The generic programming facilities of the C++ language aid in the development of general purpose reusable components.
The choice to declare \verb+ioa::automaton+ as a virtual base class is discussed in the paragraph on static composition.

@@

\paragraph{Declaring/defining actions.}
Actions are member variables that model the action concept.
Actions contain methods for preconditions, effects, and scheduling and traits that indicate the type of action (input, output, internal), if the action is valued and the value type, and if the action is parameterized and the parameter type.
The traits are used to guide the execution of the methods supplying values and parameters as necessary.
The traits are also used to check that input actions are bound to output actions with the same value status and type.
Bindings are checked for type compatibility at compile time using the generic programming facilities of C++.

@@

\paragraph{Scheduling.}
Automata submit actions to the scheduler using the \verb+ioa::schedule+ function.
The \verb+ioa::schedule+ function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition.
For the \verb+asynch_lcr_automaton+, the scheduling function is:
\begin{lstlisting}
  void schedule () const {
    if (send_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::send);
    }
    if (leader_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::leader);
    }
  }
\end{lstlisting}
The various scheduling functions in the \verb+asynch_lcr_automaton+ and the constructor all dispatch to this function.

\paragraph{Static composition.}
Static composition can be accomplished by inheriting from, i.e., extending, other automata.
For example, an automaton wishing to statically compose an \verb+asynch_lcr_automaton+ with other automata could declare it as a base class and provide the necessary code to dispatch to the \verb+asynch_lcr_automaton+ actions.
A problem arises when an automaton extends more than one automaton since the common \verb+ioa::automaton+ base class will be duplicated for each inheritance path.
The solution is to make the \verb+ioa::automaton+ base class a virtual base class which eliminates the duplicates and resolves all references appropriately.

\subsection{Dynamics}

@@

\paragraph{Leader election in a ring using TCP.}
The \verb+asynch_lcr_automaton+ is a reusable component designed to be used in the context of a larger system.
The \verb+tcp_lcr_automaton+ listed in \verb+examples/tcp_lcr.cpp+ elects a leader in a dynamic ring of processes connected by TCP sockets.
The structure of the \verb+tcp_lcr_automaton+ is given in Figure~\ref{tcp_lcr_automaton}.
The \verb+tcp_lcr_automaton+ is composed with a \verb+tcp_acceptor_automaton+ to accept a connection from its predecessor, a \verb+tcp_connector_automaton+ to create a connection to its successor, and two \verb+tcp_connection_automaton+ representing the predecessor and successor, and a \verb+asynch_lcr_automaton+ to realize the leader election algorithm.
The leader of the ring periodically sends a token which is forwarded by all nodes in the ring.
If a node has not received a token for a specified duration of time, it calls for an election.

@@

\paragraph{Dynamic composition.}
Dynamic composition is accomplished by creating child automata and bindings.
The \verb+tcp_lcr_automaton+ creates a \verb+asynch_lcr_automaton+ and binds to it with the following code:
\begin{lstlisting}
    ioa::automaton_manager<asynch_lcr_automaton<uuid> >* lcr =
       ioa::make_automaton_manager (this,
         ioa::make_generator<asynch_lcr_automaton<uuid> > (m_u));
    ioa::make_binding_manager (this,
			       lcr, &asynch_lcr_automaton<uuid>::send,
			       &m_self, &tcp_lcr_automaton::send_lcr);
    ioa::make_binding_manager (this,
			       &m_self, &tcp_lcr_automaton::receive_lcr,
			       lcr, &asynch_lcr_automaton<uuid>::receive);
    ioa::make_binding_manager (this,
			       lcr, &asynch_lcr_automaton<uuid>::leader,
			       &m_self, &tcp_lcr_automaton::leader_lcr);
    ioa::make_binding_manager (this,
			       &m_self, &tcp_lcr_automaton::init_lcr,
			       lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
Children automata are managed by \verb+ioa::automaton_manager+ objects which can be created via the \verb+ioa::make_automaton_manager+ function.
An automaton manager object requires an \verb+ioa::automaton+ object for its system action implementation and a generator for allocating the child automaton.
Bindings are managed by \verb+ioa::binding_manager+ objects which can created via the \verb+ioa::make_binding_manager+ function.
A binding manager objects requires an \verb+ioa::automaton+ object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The \verb+ioa::automaton_manager+ class and \verb+ioa::binding_manager+ class contain methods for destroying and unbinding respectively.
The \verb+m_self+ object of the example is a \verb+ioa::handle_manager+ which can be used to bind to automata that already exist.
From the example, the \verb+tcp_lcr_automaton+ binds the \verb+asynch_lcr_automaton+ to itself.
The wary reader might notice that the \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+s in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the \verb+ioa::automaton+ and deallocated at the appropriate time.

@@

\paragraph{System action events.}
Determining the outcome of a system action event, e.g., that a child automaton was created, is necessary when there are constraints on the order in which system actions must occur.
Specifically, an automaton must be able to receive events from its \verb+ioa::automaton+ base class.
This is a straightforward application of the Observer Pattern~\cite{} where the user automaton observes events forwarded by the \verb+ioa::automaton+ base class.
Conceptually, the system input actions of the \verb+ioa::automaton+ base class are mapped into the user automaton.

@@

\paragraph{Handling errors by replacing.}
The non-deterministic and the asynchronous nature of system actions complicates replacing an automaton.
To replace a child automaton, the parent automaton must destroy the child, create a new child, and bind to the new child.
For reasons already discussed, the bind must occur after the create.
This constraint is enforced by the \verb+ioa::binding_manager+ class.
Additionally, the destroy must occur before the first bind to ensure that all bindings associated with the first child have been dissolved.
Attempts to bind before the destroy will fail because either the output action or the input action will be unavailable, i.e., still bound to the first child.
A general solution is to observe the automaton and wait for it to be destroyed.

>>>>

\paragraph{Declaring an automaton.}
All automata inherit from the common base class \verb+ioa::automaton+ which implements the system action interface.
The following code declares the \verb+asynch_lcr_automaton+:
\begin{lstlisting}
template <typename UID>
class asynch_lcr_automaton :
  public ioa::automaton
{
  ...
};
\end{lstlisting}
The \verb+asynch_lcr_automaton+ is declared as a template where the \verb+UID+ template parameter must model the unique identifier concept.
The generic programming facilities of the C++ language aid in the development of general purpose reusable components.

@@

\paragraph{Declaring/defining actions.}
Actions are member variables that model the action concept.
Actions contain methods for preconditions, effects, and scheduling and traits that indicate the actions's type (input, output, internal), value status, value type, parameter status, and parameter type.
The traits are used to guide the execution of the methods by the model and supply values and parameters as necessary.
The traits are also used to check that input actions are bound to output actions with the same value status and type.
Bindings are checked for type compatibility at compile time using the generic programming facilities of C++.

@@

\paragraph{Scheduling.}
Automata submit actions to the scheduler using the \verb+ioa::schedule+ function.
The \verb+ioa::schedule+ function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
For the \verb+asynch_lcr_automaton+, the scheduling function is:
\begin{lstlisting}
  void schedule () const {
    if (send_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::send);
    }
    if (leader_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::leader);
    }
  }
\end{lstlisting}
The various scheduling functions in the \verb+asynch_lcr_automaton+ and the constructor all dispatch to this function.

\subsection{Dynamic Composition}

@@

\paragraph{Leader election in a ring using TCP.}
The \verb+asynch_lcr_automaton+ is a reusable component designed to be used in the context of a larger system.
The \verb+tcp_lcr_automaton+ listed in \verb+examples/tcp_lcr.cpp+ elects a leader in a dynamic ring of processes connected by TCP sockets.
The structure of the \verb+tcp_lcr_automaton+ is given in Figure~\ref{tcp_lcr_automaton}.
The \verb+tcp_lcr_automaton+ is composed with a \verb+tcp_acceptor_automaton+ to accept a connection from its predecessor, a \verb+tcp_connector_automaton+ to create a connection to its successor, two \verb+tcp_connection_automaton+ representing the predecessor and successor, and a \verb+asynch_lcr_automaton+ to realize the leader election algorithm.
The leader of the ring periodically sends a token which is forwarded by all nodes in the ring.
If a node has not received a token for a specified duration of time, it calls for an election.

@@

\paragraph{Dynamic composition.}
Dynamic composition is accomplished by creating child automata and bindings via \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+ objects.
The \verb+tcp_lcr_automaton+ creates a \verb+asynch_lcr_automaton+ and binds to it with the following code:
\begin{lstlisting}
    ioa::automaton_manager<asynch_lcr_automaton<uuid> >* lcr =
       ioa::make_automaton_manager (this,
         ioa::make_generator<asynch_lcr_automaton<uuid> > (m_u));
    ioa::make_binding_manager (this,
			       lcr, &asynch_lcr_automaton<uuid>::send,
			       &m_self, &tcp_lcr_automaton::send_lcr);
    ioa::make_binding_manager (this,
			       &m_self, &tcp_lcr_automaton::receive_lcr,
			       lcr, &asynch_lcr_automaton<uuid>::receive);
    ioa::make_binding_manager (this,
			       lcr, &asynch_lcr_automaton<uuid>::leader,
			       &m_self, &tcp_lcr_automaton::leader_lcr);
    ioa::make_binding_manager (this,
			       &m_self, &tcp_lcr_automaton::init_lcr,
			       lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
Children automata are managed by \verb+ioa::automaton_manager+ objects which can be created via the \verb+ioa::make_automaton_manager+ function.
An automaton manager object requires an \verb+ioa::automaton+ object for its system action implementation and a generator for allocating the child automaton.
Bindings are managed by \verb+ioa::binding_manager+ objects which can created via the \verb+ioa::make_binding_manager+ function.
A binding manager objects requires an \verb+ioa::automaton+ object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The \verb+ioa::automaton_manager+ class and \verb+ioa::binding_manager+ class contain methods for destroying and unbinding respectively.
The \verb+m_self+ object of the example is a \verb+ioa::handle_manager+ which can be used to bind to automata that already exist.
From the example, the \verb+tcp_lcr_automaton+ binds the \verb+asynch_lcr_automaton+ to itself.
The wary reader might notice that the \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+s in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the \verb+ioa::automaton+ and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.

@@

\paragraph{System action events.}
Determining the outcome of a system action event, e.g., that a child automaton was created, is necessary when there are constraints on the order in which system actions must occur.
For example, the Binding Manager mentioned in section~\ref{system_action_section} waits until the automata supply the output action and input action are created before attempting to bind.
User automata can observe \verb+ioa::automaton_manager+ objects and \verb+ioa::binding_manager+ objects to receive events generated upon the completion of a system action.

@@

\paragraph{Handling errors by replacing.}
The non-deterministic and asynchronous nature of system actions complicates replacing an automaton.
To replace a child automaton, the parent automaton must destroy the child, create a new child, and bind to the new child.
For reasons already discussed, the bind must occur after the create.
This constraint is enforced by the \verb+ioa::binding_manager+ class.
Additionally, the destroy must occur before the first bind to ensure that all bindings associated with the first child have been dissolved.
Attempts to bind before the destroy will fail because either the output action or the input action will be unavailable, i.e., still bound to the first child.
A general solution is to observe the automaton and wait for it to be destroyed.
